Practical Optional Types for Clojure

### 2015-02-28 formal-model.tex
formatting

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    \standardsubtyping{}
  \end{mathpar}
  \caption{Core Subtyping rules}
  \label{main:figure:subtyping}
\end{figure*}

@@

\begin{figure}[h]
  \footnotesize

  $$
  \begin{altgrammar}
  \end{altgrammar}
  $$
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
            &\alt& \reflectiveexpsyntax{} \\
     &\alt& \mininonreflectiveexpsyntax{}
    \\

@@

The evaluation rules B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
arguments and call the relevant JVM operation, which do not model---\secref{sec:metatheory}
states our exact assumptions about each.

@@

The key component of the T-IsA rule is the \isacompareliteral{} 
metafunction
(\figref{main:figure:mmsyntax}), used to calculate the propositions for
\isaliteral{} tests.

\begin{mathpar}
  \TIsA{}
\end{mathpar}

@@

\begin{figure}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt {\createmultiexp {\t{}} {\e{}}} &\mbox{Expressions} \\
             &\alt& {\extendmultiexp {\e{}} {\e{}} {\e{}}}
             \alt {\isaapp {\e{}} {\e{}}}\\
  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
                &\mbox{Values} \\
  \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
                &\mbox{Types} \\

@@

To ease presentation, we present \emph{immutable}
multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. \egref{example:rep} is now written
\begin{minted}{clojure}
(let [path (defmulti [Any -> (U nil String)] class)]
  (let [path (defmethod path String [x] x)]
    (let [path (defmethod path File [^File x] 
                 (.getPath x))]
      (path "dir/a")))) ;=> "a"
\end{minted}

The syntax for multimethods is given in \figref{main:figure:mmsyntax}.
{\MultiFntype {\s{}} {\sp{}}} is the type of multimethods with \emph{interface type}
{\s{}} and \emph{dispatch function type} {\sp{}}.
The expression {\createmultiexp {\s{}} {\e{}}} expression creates a multimethod
with interface type \s{} and dispatch function \e{}.
The expression {\extendmultiexp {\e{m}} {\e{v}}{\e{f}}}
takes multimethod \e{m} and returns a new multimethod with an extended dispatch
table from dispatch value {\e{v}} to {\e{f}}.
The value {\multi {\v{}} {\disptable{}}} is the runtime value of a multimethod
with dispatch function {\v{}} and dispatch table {\disptable{}}.

The typing rules are in \figref{main:figure:mmsyntax}.
The T-DefMulti rule checks to ensure the type of dispatch function
is a function that has at least as permissive a parameter type
as the interface type.

For example, let us check the \defmultiliteral{} from our translation of \egref{example:rep}
using T-DefMulti.

$$
\judgement{\propenv{}}
{\hastype
{\createmultiexp 
      {\s{}}
      {\classconst{}}}
  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
where \s{}  = {\ArrowOne {\x{}} {\Top{}} {\t{}} {\filterset {\topprop{}} {\topprop{}}} {\emptyobject{}}}
  and \sp{} = {\ArrowOne {\x{}} {\Top{}} {\Union{\Nil}{\Class}} {\filterset {\topprop{}} {\topprop{}}} {\path{\classpe{}}{\x{}}}}.
  Since the parameter types agree, this is well-typed.

The T-DefMethod rule is carefully constructed to ensure we have a syntactic
lambda expression as the right-most subexpression.
This way we can manually check the body of the lambda under an extended
environment like we sketched in \egref{example:incmap}.
We use \isacompareliteral{} to infer the proposition for this method.

Let us continue with the next line of the translation of \egref{example:rep}.
From our previous line we have \propenvp{} = \propenv{},{\isprop{\MultiFntype {\s{}}{\sp{}}}{path}}.
$$
\judgement{\propenvp{}}
{\hastype
  {\extendmultiexp {prop} {\String}
                   {\abs {\x{}} {\Top{}} {\x{}}}}
  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
We know \emph{prop} is a multimethod by \propenvp{}, so now we check the body.
$$
\judgement{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
{\hastype
  {\x{}}
  {\String}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
This is checked by T-Local since {\inpropenv{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
The new proposition {\isprop{\String}{\x{}}} was inferred by 
$$
  \isacompare{\Top{}}{\path{\classpe{}}{\x{}}}{\Value{\File{}}}
             {\filterset{\isprop{\String}{\x{}}}
                        {\notprop{\String}{\x{}}}}
$$

The next line is similarly. The final line is checked by T-App because
subtyping for multimethods (\figref{main:figure:mmsyntax}), specifically S-PMultiFn
says that {\MultiFntype {\s{}}{\sp{}}} can be upcast to its interface type {\s{}}.
Since {\s{}} has parameter type \Top{}, this holds trivially.

Multimethod semantics are in \figref{main:figure:mmsyntax}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
and uses \getmethodliteral{} (which models Clojure's \clj{get-method}) 
to obtain a method. The call to \getmethodliteral{} only returns a value
if there is exactly one method such that the corresponding dispatch value
is compatible with the current result of the dispatch function by \isaopsemliteral{}.
Finally we return the result of applying the obtained method and the original evaluated argument.

>>>>

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \SRefl{}

    \STop{}

\SUnionSuper{}

\SUnionSub{}

\SFunMono{}

\SObject{}

\SClass{}

\SSBool{}

\SFun{}

\SSKw{}

  \end{mathpar}
  \caption{Core Subtyping rules}
  \label{main:figure:subtyping}
\end{figure}

@@

\begin{figure}[t]
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
            &\alt& \reflectiveexpsyntax{} \\
     &\alt& \mininonreflectiveexpsyntax{}
    \\

@@

The evaluation rules B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
arguments and call the relevant JVM operation, which do not model---\secref{sec:metatheory}
states our exact assumptions.

@@

The key component of the T-IsA rule is the \isacompareliteral{} 
metafunction
(\figref{main:figure:mmsyntax}), used to calculate the propositions for
\isaliteral{} tests.
\begin{mathpar}
  \TIsA{}
\end{mathpar}

@@

\begin{figure}[t!]
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt {\createmultiexp {\t{}} {\e{}}} &\mbox{Expressions} \\
             &\alt& {\extendmultiexp {\e{}} {\e{}} {\e{}}}
             \alt {\isaapp {\e{}} {\e{}}}\\
  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
                &\mbox{Values} \\
  \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
                &\mbox{Types} \\

@@

To ease presentation, we present \emph{immutable}
multimethods, with syntax and semantics given in \figref{main:figure:mmsyntax}. 
\defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. \egref{example:rep} is now written
\begin{minted}{clojure}
(let [path (defmulti [Any -> (U nil String)] class)]
  (let [path (defmethod path String [x] x)]
    (let [path (defmethod path File [^File x] 
                 (.getPath x))]
      (path "dir/a")))) ;=> "a"
\end{minted}

The type {\MultiFntype {\s{}} {\sp{}}} characterizes multimethods with \emph{interface type}
{\s{}} and \emph{dispatch function type} {\sp{}}.
The expression {\createmultiexp {\s{}} {\e{}}} defines a multimethod
with interface type \s{} and dispatch function \e{}.
The expression {\extendmultiexp {\e{m}} {\e{v}}{\e{f}}}
extends multimethod \e{m} and to map
dispatch value {\e{v}} to {\e{f}} in an extended dispatch table.
The value {\multi {\v{}} {\disptable{}}} is the runtime value of a multimethod
with dispatch function {\v{}} and dispatch table {\disptable{}}.

The T-DefMulti rule ensures that the type of the dispatch function
has at least as permissive a parameter type
as the interface type.
%
For example, we can check the definition from our translation above of \egref{example:rep}
using T-DefMulti.
$$
\judgement{}%{\propenv{}}
{\hastype
{\createmultiexp 
      {\s{}}
      {\classconst{}}}
  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
where \s{}  = {\ArrowOne {\x{}} {\Top{}} {\t{}} {\filterset {\topprop{}} {\topprop{}}} {\emptyobject{}}}
  and \sp{} = {\ArrowOne {\x{}} {\Top{}} {\Union{\Nil}{\Class}} {\filterset {\topprop{}} {\topprop{}}} {\path{\classpe{}}{\x{}}}}.
  Since the parameter types agree, this is well-typed.

The T-DefMethod rule is carefully constructed to ensure we have a syntactic
lambda expression as the right-most subexpression.
This way we can manually check the body of the lambda under an extended
environment as sketched in \egref{example:incmap}.
We use \isacompareliteral{} to compute the proposition for this method,
since \isaliteral{} is used at runtime in multimethod dispatch.

We continue with the next line of the translation of \egref{example:rep}.
From the previous line we have \propenv{} = {\isprop{\MultiFntype {\s{}}{\sp{}}}{path}},
so
$$
\judgement{\propenv{}}
{\hastype
  {\extendmultiexp {prop} {\String}
                   {\abs {\x{}} {\Top{}} {\x{}}}}
  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
We know \emph{prop} is a multimethod by \propenv{}, so now we check the body
of this method.
$$
\judgement{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
{\hastype
  {\x{}}
  {\String}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
%This is checked by T-Local since {\inpropenv{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
The new proposition {\isprop{\String}{\x{}}} is derived by 
$$
  \isacompare{\Top{}}{\path{\classpe{}}{\x{}}}{\Value{\File{}}}
             {\filterset{\isprop{\String}{\x{}}}
                        {\notprop{\String}{\x{}}}}.
$$
%
The body of the \clj{let} is checked by T-App because
{\MultiFntype {\s{}}{\sp{}}} is a subtype of its interface type {\s{}}.

Multimethod definition semantics are straightforward.
B-DefMulti creates a multimethod with the given dispatch function and an empty dispatch table.
B-DefMethod produces a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to obtain the dispatch value,
and uses \getmethodliteral{} (which models Clojure's \clj{get-method}) 
to extract the appropriate method. The call to \getmethodliteral{} only returns a value
if there is \emph{exactly one} method such that the corresponding dispatch value
is compatible, using \isaopsemliteral{}, with the result of the dispatch function.
Finally we return the result of applying the extracted method and the original argument.
